Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-24 BN_discretizer_paper.tex
Do all todos
Change the description of Lustgarten's paper in the introduction

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
  gsave
  newpath
    20 20 moveto
    20 220 lineto
    220 220 lineto
    220 20 lineto
  closepath
  2 setlinewidth
  gsave
    .4 setgray fill
  grestore
  stroke
  grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{groupplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{boondox-cal} % for lowercase caligraphic
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\pgfplotsset{
    jitter/.style={
        x filter/.code={\pgfmathparse{\pgfmathresult+rnd*#1}}
    },
    jitter/.default=0.1
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM COMMANDS

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example are Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions and non-linear functions \citep{Ihler_2009}, these redesigned algorithms are computationally expensive and do not perform well.
The second is to use nonparametric distributions, such as particle representations and Gaussian processes \citep{Ickstadt_2010}.
Unlike parametric methods, nonparametric methods do not make assumptions about the probability distributions being used.
Parametric models have a fixed number of parameters, whereas the number of parameters in a nonparametric model grow with the amount of training data.

@@

The MDL principle was first proposed by \cite{MDL_1978} and states that the best model for a dataset is one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
Results in this work suggest that the MDL method suffers from \todo{fill this out.}

@@

The MODL method uses dynamic programming to find the optimal discretization policy, and has an $\bigo\paren{n^3 + r \cdot n^2}$ runtime, where $r$ is the number of instances for the class variable.
\cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$.
Under a uniform prior probability over discretizations the runtime can be reduced to $\bigo\paren{r \cdot n^2}$ without sacrificing optimality.

@@

\begin{equation}
  \label{eq:p_D_given_M}
  P\paren{D_{-\cX} \mid \Lambda} \propto P\paren{D_{\Pa_\cX} \mid \Lambda} \cdot \prod_{C_i} P\paren{D_{C_i} \mid \Lambda, D_{\bm{S}_i}}\text{.}
\end{equation}

@@

\todo{
\begin{equation}
  \ln p\paren{D\mid \discset, D^*} = \sum_{i=1}^n
  1\curly{X_i \text{ orig. cont.}}
  \sum_{j=1}^{q_i} \sum_{k=1}^{r_i} m_{ijk} \ln\paren{
    \frac{
      1
    }{
      e^{\Lambda_{\cX_i}}_{k+1} - e^{\Lambda_{\cX_i}}_{k}
    }
  }\text{.}
\end{equation}
}

\noindent
The mean cross-validated log-likelihood is the mean log-likelihood on the witheld dataset among cross-validation folds.
Cross-validation directly estimates generalization error.
Ten folds were used in each experiment.

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
%  gsave
%  newpath
%    20 20 moveto
%    20 220 lineto
%    220 220 lineto
%    220 20 lineto
%  closepath
%  2 setlinewidth
%  gsave
%    .4 setgray fill
%  grestore
%  stroke
%  grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepgfplotslibrary{groupplots}
\usepackage{tikz}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{boondox-cal} % for lowercase caligraphic
% \usepackage[capitalize]{cleveref}
\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\pgfplotsset{
    jitter/.style={
        x filter/.code={\pgfmathparse{\pgfmathresult+rnd*#1}}
    },
    jitter/.default=0.1
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM COMMANDS

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning and inference algorithms based on their parameters.
One successful example are Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions and non-linear functions \citep{Ihler_2009}, these redesigned algorithms are computationally expensive and do not perform well.
The second is to use nonparametric distributions, such as particle representations and Gaussian processes \citep{Ickstadt_2010}.
Unlike parametric methods, nonparametric methods do not make assumptions about the probability distributions being used.
Parametric models have a fixed number of parameters, whereas the number of parameters in a nonparametric model grow with the amount of training data.

@@

The MDL principle was first proposed by \cite{MDL_1978} and states that the best model for a dataset is one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
Results in this work suggest that the MDL method suffers from low sensitivity to discretization edges and returns low numbers of discretization intervals on continuous variables.

@@

The MODL method uses dynamic programming to find the optimal discretization policy, and has an $\bigo\paren{n^3 + r \cdot n^2}$ runtime, where $r$ is the number of instances for the class variable.
\cite{Lustgarten_2011} suggest several formulations for the prior term $P(\text{Model})$.
Under the suggested prior, which includes the uniform prior probabilities in MODL as a special case, the runtime can be reduced to $\bigo\paren{r \cdot n^2}$ without sacrificing optimality.

@@

\begin{equation}
  \label{eq:p_D_given_M}
  P\paren{D_{-\cX} \mid \Lambda} \propto P\paren{D_{\Pa_\cX} \mid \Lambda} \cdot \prod_{i} P\paren{D_{C_i} \mid \Lambda, D_{\bm{S}_i}}\text{.}
\end{equation}

@@

\begin{equation}
  \ln p\paren{D\mid \discset, D^*} = \sum_{i=1}^N
  1_{\curly{ \cX_i \in \bm{\cX }}}\cdot \sum_{k=1}^{r_i}  \sum_{j=1}^{q_i} \beta_{ij}^{(k)} \ln \paren{
    \frac{
      1
    }{
      e^{\Lambda_{\cX_i}}_{k} - e^{\Lambda_{\cX_i}}_{k-1}
    }
  }\text{.}
\end{equation}
Note that $e^{\Lambda_{\cX_i}}_{0} = \textsc{Min}\paren{D_{\cX_i}}$ and $e^{\Lambda_{\cX_i}}_{r_i} = \textsc{Max}\paren{D_{\cX_i}}$.
%, where $\brock{ e^{\Lambda_{\cX_i}}_{k-1} , e^{\Lambda_{\cX_i}}_{k} }$ is the discretization interval that $\cx_i$ falls in. If $\cx_i < e^{\Lambda_{\cX_i}}_{1}$, then the corresponding interval is $\brock{ \textsc{Min}\paren{D_{X_i}} , e^{\Lambda_{\cX_i}}_{1}}$. If $\cx_i \leq e^{\Lambda_{\cX_i}}_{r_i - 1}$, then the corresponding interval is $\brock{  e^{\Lambda_{\cX_i}}_{r_i - 1},\textsc{Max}\paren{D_{X_i}}}$.
\noindent
The mean cross-validated log-likelihood is the mean log-likelihood on the witheld dataset among cross-validation folds.
Cross-validation directly estimates generalization error.
Ten folds were used in each experiment.
